---
title: "TITLE HERE"
subtitle: "Group Exercise 1 Part B"
author: "Ava Nadolski and Erik Marcinek"
date: "10/06/2026"
output:
  pdf_document:
    highlight: kate
keep_tex: no
number_sections: no
html_document:
  df_print: paged
geometry: margin = 0.5in
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
editor_options:
  chunk_output_type: console
documentclass: article
urlcolor: blue
---
  
<!-- The above is set to automatically compile to a .pdf file.   -->
<!-- It will only succeed if LaTeX is installed. -->


```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this here

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, # Recommended
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE, # Turn off warnings
                      R.options = list(width = 100))

```

\clearpage

```{r include=FALSE}
# You should not echo this chunk.
# include=FALSE does more than echo=FALSE, it actually does: echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.show='hide'

# You should set your working directory at the very beginning of your R Markdown file
# setwd("~/Dropbox/teaching/FA2020/CMDA_4654/homework/homework1/")

# In linux ~/ is shorthand for /home/username/
# You should type things out properly for your system
# Mac: /Users/username/Documents/CMDA4654/Lectures/Lecture_03/.../
# Windows: C:/Users/username/Documents/etc/Lecture/Lecture_03/.../


```

<!-- ---------------------------------------------------------------------------------------------------- -->
<!-- ---------------- Homework Problems start below these lines ----------------------------------------- -->
<!-- ---------------------------------------------------------------------------------------------------- -->

# 1. Introduction
<!-- 
briefly state your names, your assigned group number, and the chosen domain area.
---> 

# 2. Executive Summary & Research Questions
<!-- 
summarize your chosen target variables, define your explicit research questions, and explain how Part A EDA informed your variable selection.
---> 
## 2.a: Research Question


## 2.b: Classification Research Question


# 3. Methodology & Model Setup
<!-- 
detail the train/test split, data preprocessing, scaling protocols, handling of categorical levels, the probability outputs of each classifier (including any smoothing), and the threshold τ (binary) or argmax rule (multi-class).
---> 
## 3.a: Data Preprocessign and Leakage Prevention


## 3.b: Hyperparameter Optimization of kNN (Regression and Classification)


## 3.c: Naive Bayes Classification Rigor


# 4. Analysis, Tuning & Results
<!-- 
* Include tuning plots and tables for the optimal k selection (regression and classification).
* Provide the comparative evaluation table of classification metrics described in Section 4 (both models plus the no-skill baseline), with a sentence justifying your choice of metrics.
* Include at least one visualization contrasting the behavior of Naive Bayes and kNN, e.g., ROC curves overlaid on the same axes (binary), a calibration plot, or a 2D decision boundary plot.
---> 
## 4.1: Regression (kNN Regression)

**Root Mean Squared Error (RMSE):**
**Mean Absolute Error (MAE):**
**Coefficient of Determination** ($R^2$)**:**

## 4.2 Classification (kNN & Naive Bayes)

**Confusion Matrices**
**Log-loss from the predicted probabilites**
* Binary targets
* Multi-class targets

**Null Baseline Entropy**

At least three more metrics of your choice
Choose at least three additional metrics, including at least one from each group:

# 5. Discussion & Real-World Implications
<!-- 
address the discussion prompts listed in Section 6.
--->

# 6. Citations & Data Link
<!-- 
link to the original data source and provide a direct, one-click access link to your data file (e.g., hosted on Google Drive).
--->
Baseball Savant Website: [Original data source](https://baseballsavant.mlb.com/statcast_search)
Data File: [Clean Data](https://docs.google.com/spreadsheets/d/1_zxPBT3iTcbFs3PdsiJ9JdB-IF8XxnXb9x7-mVuE-XU/edit?usp=sharing)

# 7. Appendix
<!-- 
all clean, commented R code necessary to fully reproduce your training splits, models, metric calculations, tables, and figures.
--->

