---
title: "Group 3 Exercise 1 Finale"
author: "Ava Nadolski and Erik Marcinek"
date: "2026-10-08"
fontsize: 9pt
classoption: "aspectratio=169"
output: 
  beamer_presentation: 
    fonttheme: professionalfonts
    highlight: kate
    theme: Rochester
header-includes:
- \definecolor{VTmaroon}{HTML}{861F41}
- \usecolortheme[named=VTmaroon]{structure}
- \usepackage{tikz}
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})

```

## Introduction & Research Questions
**Data**

- Play-by-play from Baseball Savant. 
- 25000 observations
- 100 categorical/numerical variables

**Research Questions**

- **Regression:** Can we predict how far a ball will go given predictors launch speed/angle, attack angle/direction, swing path tilt, and horizontal distance between batter's center of mass and contact point.

- **Classification:** Can we accurately predict a pitch result (ball/strike/in play) from pre-pitch information such as pitch type, ball-strike count, velocity, spin rate, horizontal/vertical break, and pitcher/batter handedness. And how does it compare from kNN and Naive Bayes?
  -Why?: Pitchers, catchers, and defenses could anticipate what a pitch will produce.

## Regression: Exploratory Data Analysis (EDA)

Visual Storytelling: Present key summary visualizations or tables from your EDA that directly motivate or support your research questions.
Insights: Highlight any interesting patterns, anomalies, or distributions discovered prior to modeling.


## Regression: Data Science Methods & Techniques

**Methodology**

**Justification**


## Regression: Results & Key Findings




## Classification: Exploratory Data Analysis (EDA)
::::{.columns}
:::{.column width="35%"}

 \hfill\break
  \hfill\break
   \hfill\break
    \hfill\break
     \hfill\break
      \hfill\break

- Pitch type influences strike percentage.
```{r, echo=FALSE}
baseballData <- read.csv("/Users/avanadolski/Desktop/CMDA4654/Group HW 1 Part A/savant_data.csv")
library(ggplot2)
```
:::
:::{.column width="65%"}
```{r, echo = FALSE, fig.height=2.8, fig.width=3.7, fig.align='right'}
# remove null values
baseballData_clean <- baseballData[!is.na(baseballData$pitch_name) & baseballData$pitch_name != "", ]
# aggregate pitch name into strike percentage categories
pitch_summary <- aggregate(
type ~ pitch_name,
data = baseballData_clean,
# calculating strike percentage
FUN = function(x) c(n = length(x), strike_pct = mean(x %in% c("S", "X")) * 100))
# put all values into a new data frame
pitch_summary <- data.frame(
pitch_name = pitch_summary$pitch_name,
n_pitches = pitch_summary$type[, "n"],
strike_pct = pitch_summary$type[, "strike_pct"]
)
# remove pitches that are rare
pitch_summary <- pitch_summary[pitch_summary$n_pitches >= 10, ]
# create plot
ggplot(pitch_summary, aes(x = reorder(pitch_name, strike_pct), y = strike_pct)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(strike_pct, 1)), hjust = -0.1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  coord_flip() +
  labs(
    title = "Strike Percentage by Pitch Type",
    x = "Pitch Type",
    y = "Strike %"
  ) +
  theme_minimal(base_size = 9)+
  theme(plot.title.position = "plot")
```
:::
::::

## Classification: Exploratory Data Analysis (EDA)
::::{.columns}
:::{.column width="35%"}

 \hfill\break
  \hfill\break
   \hfill\break
    \hfill\break
     \hfill\break
      \hfill\break

- Influence of ball-strike count on pitch types.

:::
:::{.column width="65%"}
```{r, echo = FALSE, fig.height=2.8, fig.width=3.7, fig.align='right'}
# define more general pitch type groupings
fastballs <- c("4-Seam Fastball", "Sinker", "Cutter", "Fastball")
breaking <- c("Slider", "Curveball", "Sweeper", "Slurve", "Knuckle Curve", "Screwball", "Slow Curve")
offspeed <- c("Changeup", "Splitter", "Forkball", "Eephus", "Knuckleball", "Slurve")
# group pitch types based on the above groupings
pitch_group <- ifelse(
baseballData$pitch_name %in% fastballs, "Fastball",
ifelse(baseballData$pitch_name %in% breaking, "Breaking Ball",
ifelse(baseballData$pitch_name %in% offspeed, "Offspeed", "Other"))
)
# combine balls and strikes into one value
count <- paste(baseballData$balls, baseballData$strikes, sep = "-")
# factor new count variable
count_order <- c("0-0","0-1","0-2","1-0","1-1","1-2","2-0","2-1","2-2","3-0","3-1","3-2")
count <- factor(count, levels = count_order)
# create plot
ggplot(baseballData, aes(x = count, fill = pitch_group)) +
geom_bar(position = "fill") +
scale_y_continuous(labels = scales::percent) +
scale_fill_manual(values = c(
"Fastball" = 'darkblue',
"Breaking Ball" = "blue",
"Offspeed" = "lightblue",
"Other" = "grey"
)) +
labs(
title = "Which Pitch is Most Common Given a Specific Count",
x = "Count (Balls-Strikes)",
y = "Proportion of Pitches",
fill = "Pitch Category"
) +
theme_minimal(base_size = 6) +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
:::
::::

## Classification: Exploratory Data Analysis (EDA)

- Training proportions heavily favor Strike. 

|Class  | Proportion|
|:------|----------:|
|Ball   |      0.354|
|Strike |      0.481|
|InPlay |      0.165|


## Classification: Data Science Methods & Techniques

**Shared setup:** 80/20 train/test split, tuning training data only, pitch types with under 50 training pitches were collapsed into "Other".

:::::: {.columns}
::: {.column width="50%"}
**kNN**

- Pitch name, pitcher hand, batter stance are dummy coded (no natural order)
- Numeric features z-scored.
- k* = 201, chosen by 5-fold CV on log-loss.
- Smoothed vote fractions give probabilities (no zeros).
:::
::: {.column width="50%"}
**Naive Bayes**

- Numeric features: Gaussian densities so no scaling needed
- Pitch name, hands, balls, strikes is treated with Laplace smoothing
:::
::::::


## Classification: Key Findings
- kNN stays near the diagonal, but only because k = 201 pulls its probabilities toward the class rates.
- Naive Bayes’ curves are flatter than the diagonal in the tails, which means it is overconfident.
  - Speed and vertical break (r = 0.72) are double-counted, violates Naive Bayes' independence.

\begin{center}
\includegraphics[width=0.75\textwidth, height=0.65\textheight]{/Users/avanadolski/Desktop/calibrationplots.png}
\end{center}

## Classification: Key Findings
Confusion matrices. Rows = predicted, columns = actual.


:::::: {.columns}
::: {.column width="50%"}



| **kNN**  | Ball| Strike| InPlay|
|:------|----:|------:|------:|
|Ball   |  312|    305|    139|
|Strike | 1503|   2025|    703|
|InPlay |    0|      0|      0|

:::
::: {.column width="50%"}



|**Naive Bayes**| Ball| Strike| InPlay|
|:------|----:|------:|------:|
|Ball   |  535|    599|    219|
|Strike | 1257|   1705|    609|
|InPlay |   23|     26|     14|

:::
::::::
- Both models most accurately predict Strike.
- For kNN, InPlay is never predicted. Likely because the no-skill baseline is 0.165.

## Classification: Key Findings


|Metric                         | No-skill baseline|    kNN| Naive Bayes|Better model |
|:------------------------------|-----------------:|------:|-----------:|:------------|
|Accuracy                       |            0.4672| 0.4686|      0.4520|kNN          |
|Balanced accuracy              |            0.3333| 0.3470|      0.3477|Naive Bayes  |
|MCC                            |            0.0000| 0.0427|      0.0327|kNN          |
|Log-loss                       |            1.0241| 1.0133|      1.0413|kNN          |
|Log-loss improvement over null |            0.0000| 0.0105|     -0.0168|kNN          |
|Brier score                    |            0.6211| 0.6166|      0.6344|kNN          |

  
- kNN wins for all except balanced accuracy, but not by much. Therefore, **kNN predicts  better**, but only at a slight margin.


## Impact & Discussion


**Regression:**

**Classification:**

- Costly misses: Missing a ball hurts the defense the most, both models do a poor job at predicting them.
- Remove double counted variables: Release speed and vertical break are strongly correlated, so Naive Bayes is counting same information twice.
- Need better/more predictor variables: Spin, speed, and break carry little signal.  





