---
title: "Group Exercise 1"
subtitle: "Part B"
author: "Ava Nadolski and Erik Marcinek"
date: "10/06/2026"
output:
  pdf_document:
    highlight: kate
keep_tex: no
number_sections: no
html_document:
  df_print: paged
geometry: margin = 0.5in
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
editor_options:
  chunk_output_type: console
documentclass: article
urlcolor: blue
---
  
<!-- The above is set to automatically compile to a .pdf file.   -->
<!-- It will only succeed if LaTeX is installed. -->


```{r setup, include=FALSE}
# This is the setup chunk
#  Here you can set global options for the entire document

library(knitr) # I recommend doing this here

# Although you can call functions from a library using the following notation
#  without loading the entire library.
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA, # Recommended
                      fig.path = "./figures/",  # Store all figures here in relative path (make the folder first)
                      fig.align = "center",
                      fig.width = 7,
                      fig.height = 7,
                      message = FALSE, # Turn off load messages
                      warning = FALSE, # Turn off warnings
                      R.options = list(width = 100))

```

\clearpage

```{r include=FALSE}
# You should not echo this chunk.
# include=FALSE does more than echo=FALSE, it actually does: echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.show='hide'

# You should set your working directory at the very beginning of your R Markdown file
# setwd("~/Dropbox/teaching/FA2020/CMDA_4654/homework/homework1/")

# In linux ~/ is shorthand for /home/username/
# You should type things out properly for your system
# Mac: /Users/username/Documents/CMDA4654/Lectures/Lecture_03/.../
# Windows: C:/Users/username/Documents/etc/Lecture/Lecture_03/.../


```

<!-- ---------------------------------------------------------------------------------------------------- -->
<!-- ---------------- Homework Problems start below these lines ----------------------------------------- -->
<!-- ---------------------------------------------------------------------------------------------------- -->

# 1. Introduction
<!-- 
briefly state your names, your assigned group number, and the chosen domain area.
---> 
This report was prepared by Ava Nadolski and Erik Marcinek (Group 3). Our chosen domain is sports, more specifically play-by-play baseball data from Baseball Savant. In Part A we contrusted an exploratory data analysis (EDA) of the data. In Part B we build, tune, and evaluate regression and classification models through kNN and Naive Bayes.

# 2. Executive Summary & Research Questions
<!-- 
summarize your chosen target variables, define your explicit research questions, and explain how Part A EDA informed your variable selection.
---> 
## 2.a: Research Question


## 2.b: Classification Research Question
Which classifier, kNN or Naive Bayes, can more accurately predict a pitch result (ball/strike/in play) from pre-pitch information such as pitch type, ball-strike count, velocity, spin rate, horizontal/vertical break, and pitcher/batter handedness?

The target is the Statcast pitch result `type`, with $K = 3$ classes: `Ball` (B), `Strike` (S), and `InPlay` (X). Predicting pitch results matters because it lets pitchers, catchers, and defenses anticipate what a given pitch is likely to produce. Below is a table with all variables and a connection to the EDA done previously. All plot names refer to the EDA from Nadolski_Group_Exercise1.

|Variable|Type|Justification|
|--|--|-----|
|Pitch Type| Categorical|Plot 1: Strike percentage differ by pitch type|
|Balls| Numerical|Plot 3: Pitch type differs by ball-strike count.|
|Strikes| Numerical|Plot 3: Pitch type differs by ball-strike count.|
|Velocity| Numerical|Plot 4: Velocity differs by pitch outcome (Ball, Foul, other).|
|Spin Rate| Numerical|Table 2: Spin differs by pitch type.|
|Horizontal Break| Numerical|Table 2: Horizontal break differs by pitch type.|
|Vertical Break| Numerical|Table 2: Vertical break differs by pitch type.|
|Pitcher Handedness| Categorical| Domain knowledge, same batter-handedness advantage.|
|Batter Handedness| Categorical| Domain knowledge, opposite pitcher-handedness advantage.|

 \hfill\break

```{r, echo=FALSE}
# Load data and libraries
baseballData <- read.csv("/Users/avanadolski/Desktop/CMDA4654/Group HW 1 Part A/savant_data.csv")

library(knitr)
library(tidyverse)
library(FNN)
library(e1071) 
library(ggplot2)
library(patchwork)
```


```{r, echo=FALSE}
# Get subdata
classes <- c("Ball", "Strike", "InPlay")
classification_data <- baseballData %>%
  transmute(Y = factor(type, levels = c("B", "S", "X"), labels = classes),
            pitch_name = as.character(pitch_name),
            balls, strikes, release_speed, release_spin_rate,
            pfx_x = abs(pfx_x), pfx_z,
            stand = as.character(stand), p_throws = as.character(p_throws))
```


```{r, echo=FALSE}
# Handle missingness delibrately 
# Keeps only the rows where 'Y' is not NA
n_raw <- nrow(classification_data)
classification_data <- classification_data[complete.cases(classification_data) &
                                           classification_data$pitch_name != "", ]
n_removed <- n_raw - nrow(classification_data)
```


```{r, echo=FALSE}
set.seed(1234)
# Split data into 80%/20% training/testing split
index <- sample(1:nrow(classification_data), round(nrow(classification_data) * 0.8))
training_df <- classification_data[index, ]
testing_df <- classification_data[-index, ]
```


```{r, echo=FALSE}
# training class proportions
K <- length(classes)

pi_train <- prop.table(table(training_df$Y))
kable(data.frame(Class = classes, Proportion = round(as.numeric(pi_train), 3)),
      caption = "Training class proportions $\\pi_k$.")
```


```{r, echo=FALSE}
# No-skill model baselines
majority <- names(which.max(pi_train))
null_acc <- mean(testing_df$Y == majority)                              
null_ll <- -mean(log(pi_train[as.character(testing_df$Y)]))
null_ll_train <- -sum(pi_train * log(pi_train))

baselines <- c(accuracy_NIR = max(pi_train),
               balanced_acc = 1 / K,
               MCC = 0, kappa = 0,
               log_loss = null_ll_train,
               brier = 1 - sum(pi_train^2))

baseline_tab <- data.frame(
  Metric = c("Accuracy", "Balanced accuracy", "MCC", "Cohen's kappa",
             "Log-loss", "Brier score"),
  `Training proportions` = round(baselines, 3),
  `Test set` = c(round(null_acc, 3), "—", "—", "—", round(null_ll, 3), "—"),
  check.names = FALSE)

kable(baseline_tab, row.names = FALSE,
      caption = "No-skill baselines")
```


# 3. Methodology & Model Setup
<!-- 
detail the train/test split, data preprocessing, scaling protocols, handling of categorical levels, the probability outputs of each classifier (including any smoothing), and the threshold τ (binary) or argmax rule (multi-class).
---> 

## 3.a: Data Preprocessing and Leakage Prevention
**Classification**
**Data split:** 
We used an 80/20 train/test split with `set.seed(1234)` and the test set was touched only once, after all tuning was finished.

```{r, echo=FALSE}
# Combine pitches with <50 training pitches into 'Other' (levels chosen from TRAIN only)
keep <- names(which(table(training_df$pitch_name) >= 50))
collapse_pitch <- function(x) ifelse(x %in% keep, x, "Other")
training_df$pitch_name <- collapse_pitch(training_df$pitch_name)
testing_df$pitch_name  <- collapse_pitch(testing_df$pitch_name)
```

**Rare pitch types:** 
From the EDA, many pitch types appeared very few times. For example, forkball and pitch out have only one observation each. Categories that small make estimates unstable for both models, so pitch types with fewer than 50 training observations were collapsed into an "Other" category. The collapsing rule was performed after the split, so the test set did not influence it.

```{r, echo=FALSE}
# turn categorical vars into factors
cat_vars <- c("pitch_name", "stand", "p_throws")
for (v in cat_vars) {
  lv <- sort(unique(training_df[[v]]))
  training_df[[v]] <- factor(training_df[[v]], levels = lv)
  testing_df[[v]]  <- factor(testing_df[[v]],  levels = lv)
}
```

**kNN Standardization:** 

```{r, echo=FALSE, results='hide', fig.height=4, fig.width=7, fig.align="center"}
cont_vars <- c("release_speed", "release_spin_rate", "pfx_x", "pfx_z") 

long <- training_df %>%
  select(Y, all_of(cont_vars)) %>%
  pivot_longer(-Y, names_to = "feature", values_to = "value")

# Histograms
hist <- ggplot(long, aes(value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, fill = "grey80", colour = "white") +
  geom_density(colour = "blue") +
  facet_wrap(~ feature, scales = "free") +
  labs(title = "Distribution of each cont. feature")

# Boxplots 
box <- ggplot(long, aes(x = feature, y = value)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~ feature, scales = "free") +
  labs(title = "Boxplots of each cont. feature", x = NULL)

hist + box
```

```{r, echo=FALSE, results='hide', fig.height=3, fig.width=4, fig.align="center"}
# QQ plots
qqplots <- ggplot(long, aes(sample = value)) +
  stat_qq(alpha = 0.3) + stat_qq_line(colour = "red") +
  facet_wrap(~ feature, scales = "free") +
  labs(title = "Normal Q-Q plots of each cont. feature")
qqplots
```

```{r, echo=FALSE}
# standardize by z-score

numerical_vars <- c("balls", "strikes", "release_speed", "release_spin_rate", "pfx_x", "pfx_z")
train_scaled <- scale(training_df[, numerical_vars])
center_vals <- attr(train_scaled, "scaled:center")
scale_vals  <- attr(train_scaled, "scaled:scale")
test_scaled <- scale(testing_df[, numerical_vars], center = center_vals, scale = scale_vals)

# Summary table
scale_check <- rbind(
  "Train mean" = colMeans(train_scaled),
  "Test mean"  = colMeans(test_scaled))

kable(round(scale_check, 3),
      caption = "Means in the training and test sets. Scaling parameters come from the training set only.")
```

After looking at histograms, boxplots, and qqplots, there are a few outliers in a feature release_spin_rate. Therefore, we will standardize by z-score as it is less sensitive to extreme values.

In the table above, the training set feature means are all close to zero, which is what we want. Further, our testing means are near zero, but not exactly zero. We want this too, as the test set was not used to define "zero." 

```{r, echo=FALSE, results='hide'}
# check is.ordered() to decide nominal/ordinal 
is.ordered(classification_data$stand)
is.ordered(classification_data$p_throws)
is.ordered(classification_data$pitch_name)
```

**Categorical Encoding: kNN encoding:**
Categorical variables pitch_name, stand, and p_throws have no order. There is no ranking among the pitches slider, sinker, changeup, etc. Additionally, there is no order in whether a pitcher is a righty or a lefty. Therefore, they are nominal variables and get dummy encoded.  

Pitch name, batter side, and pitcher hand are dummy coded (1/2) and left unscaled. Balls, strikes, velocity, spin rate, and horizontal/vertical break are z-scored using the training mean and SD, and those same training values are applied to the test set. Inside cross-validation, the mean and SD come from the fitting folds only.

**Categorical Encoding: Naive Bayes encoding:** 
Pitch name, batter side, pitcher hand, balls, and strikes are treated as categorical features with class-conditional frequency tables (Laplace smoothing, $\alpha = 1$). Velocity, spin rate, and horizontal/vertical break use Gaussian class-conditional densities, which is scale invariant. These features are left unstandardized.

```{r, echo=FALSE, results='hide'}
# encode nominal variables with full one-hot 
one_hot <- function(df) {
  ctr <- lapply(Filter(is.factor, df), function(f) contrasts(f, contrasts = FALSE))
  model.matrix(~ . - 1, data = df, contrasts.arg = ctr)
}

X_train <- cbind(train_scaled, one_hot(training_df[cat_vars]))
X_test <- cbind(test_scaled,  one_hot(testing_df[cat_vars]))
ytr <- as.integer(training_df$Y)

# test if variable encoding worked
stopifnot(identical(colnames(X_train), colnames(X_test)))
colnames(X_train)
```

**Probability outputs and decision rule:** 
Because the target has $K = 3$ classes, there is no single threshold $\tau$. Both models predict the class with the largest probability, $\hat{y} = \arg\max_k \hat{p}_k(x)$. kNN probabilities are smoothed vote fractions $\hat{p}_k = (\text{votes}_k + 1)/(k + K)$. Naive Bayes probabilities are the posterior class probabilities $\hat{P}(Y = k \mid x)$.The probabilities themselves are used for log-loss and the Brier score, and the argmax predictions are used for the confusion matrices and the threshold-dependent metrics.

## 3.b: Hyperparameter Optimization of kNN (Regression and Classification)
**Regression**





**Classification**
```{r, echo=FALSE}
kmax <- 301
k.grid <- c(seq(1, 51, by = 2), seq(61, kmax, by = 10))

# Class votes
knn_votes <- function(fit.X, fit.y, new.X, kmax) {
  nn <- FNN::get.knnx(fit.X, new.X, k = kmax, algorithm = "brute")$nn.index
  L  <- matrix(fit.y[nn], nrow = nrow(new.X))
  lapply(1:K, function(j) t(apply(L == j, 1, cumsum)))
}
# Laplace-smoothed vote (votes + 1) / (k + K)
smooth_p <- function(votes, k) sapply(1:K, function(j) (votes[[j]][, k] + 1) / (k + K))

multi_log_loss <- function(y, P, eps = 1e-15) {
  P <- pmin(pmax(P, eps), 1 - eps)
  -mean(log(P[cbind(seq_along(y), as.integer(y))]))
}
bal_acc <- function(y, pred) {
  tab <- table(factor(pred, 1:K), factor(y, 1:K))
  mean(diag(tab) / colSums(tab))
}
knn_metrics <- function(votes, y, k) {
  P <- smooth_p(votes, k)
  c(BalancedAcc = bal_acc(y, max.col(P, "random")), LogLoss = multi_log_loss(y, P))
}

# 5-fold CV 
set.seed(1234)
folds <- sample(rep(1:5, length.out = nrow(training_df)))

cv_knn <- function(v) {
  fit <- training_df[folds != v, ];  val <- training_df[folds == v, ]
  fit.num <- scale(fit[, numerical_vars])
  val.num <- scale(val[, numerical_vars],
                   center = attr(fit.num, "scaled:center"),
                   scale  = attr(fit.num, "scaled:scale"))
  votes <- knn_votes(cbind(fit.num, one_hot(fit[cat_vars])), as.integer(fit$Y),
                     cbind(val.num, one_hot(val[cat_vars])), kmax)
  t(sapply(k.grid, function(k) knn_metrics(votes, as.integer(val$Y), k)))
}
knn.cv <- data.frame(k = k.grid, Reduce(`+`, lapply(1:5, cv_knn)) / 5)

# Training curve on a 5,000-pitch subsample 
sub       <- sample(nrow(training_df), 5000)
votes_tr  <- knn_votes(X_train, ytr, X_train[sub, ], kmax)
knn.train <- data.frame(k = k.grid,
                        t(sapply(k.grid, function(k) knn_metrics(votes_tr, ytr[sub], k))))

# Choose k*
best_ll <- min(knn.cv$LogLoss)
k_ll <- knn.cv$k[which.min(knn.cv$LogLoss)]
k_star <- min(knn.cv$k[knn.cv$LogLoss <= best_ll + 0.001])
k_ba <- knn.cv$k[which.max(knn.cv$BalancedAcc)]
kable(data.frame(Criterion = c("Minimum CV log-loss", "Chosen k*",
                               "Maximum CV balanced accuracy"),
                 k = c(k_ll, k_star, k_ba)),
      caption = "Neighborhood sizes by each criterion")

knn.cv$LL_improvement <- 1 - knn.cv$LogLoss / null_ll_train
```

```{r, echo=FALSE, fig.width=7, fig.height=4, out.width="80%", fig.align="center"}
# Optimization plots
tune_long <- tidyr::pivot_longer(
  rbind(data.frame(set = "5-fold CV", knn.cv[, c("k", "BalancedAcc", "LogLoss")]),
        data.frame(set = "Training",  knn.train)),
  c(BalancedAcc, LogLoss), names_to = "metric", values_to = "value")
ref <- data.frame(metric = c("BalancedAcc", "LogLoss"), value = c(1 / K, null_ll_train))

ggplot(tune_long, aes(k, value, colour = set)) + geom_line() +
  geom_hline(data = ref, aes(yintercept = value), linetype = "dotted") +
  geom_vline(xintercept = k_star, linetype = "dashed") +
  scale_x_log10() + facet_wrap(~ metric, scales = "free_y") +
  labs(x = "Neighborhood size k (log scale)", y = NULL, 
       title = "Optimization Plots",
       colour = NULL)

top <- rbind(cbind(Best = "BA", knn.cv[order(-knn.cv$BalancedAcc), ][1:3, 1:3]),
             cbind(Best = "LL", knn.cv[order(knn.cv$LogLoss), ][1:3, 1:3]),
             cbind(Best = "k*", knn.cv[knn.cv$k == k_star, 1:3]))
kable(top, digits = 4, row.names = FALSE)
```

```{r, echo=FALSE}
yte <- as.integer(testing_df$Y)

votes_te <- knn_votes(X_train, ytr, X_test, k_star)
P_knn    <- smooth_p(votes_te, k_star)
colnames(P_knn) <- classes

set.seed(99)
pred_knn <- max.col(P_knn, ties.method = "random")

kable(data.frame(Metric = c("Accuracy", "Balanced accuracy", "Log-loss"),
                 Value  = round(c(mean(pred_knn == yte),
                                   bal_acc(yte, pred_knn),
                                   multi_log_loss(yte, P_knn)), 4)),
      caption = paste0("kNN test-set performance at k* = ", k_star, "."))
```

We searched neighborhood sizes with 5-fold cross-validation on the training set only. The grid started at $k \in \{1, 3, 5,..., 51\}$. The best CV $k$ landed on the boundary of 51, so the grid was extended to 301 (steps of 10 beyond 51). The log-loss minimum then landed on the new boundary of 301, but the curve is essentially flat beyond $k \approx 100$, so extending the grid further would not change the conclusion. The final $k^* =$ `r k_star` was chosen by cross-validated log-loss, using the smallest $k$ within 0.001 of the minimum because the curve is flat. The minimum log-loss was at $k =$ `r k_ll`, while balanced accuracy peaked much earlier at $k =$ `r k_ba`, so the two metrics did not agree. Balanced accuracy is nearly flat across all $k$ (roughly 0.3576 to 0.3573, against a no-skill value of `r round(1 / K, 3)`), so its differences are too small to be meaningful. Log-loss decided because it is a strictly proper scoring rule that uses the full probability vector, while balanced accuracy depends only on the argmax. Both metrics stay close to their no-skill baselines, which suggests the available pitch characteristics carry limited information about the pitch result. The test set was used exactly once, after $k^*$ was chosen, to evaluate the final kNN model.


## 3.c: Naive Bayes Classification Rigor
**Assumptions:** 
```{r, echo=FALSE}
# Naive Bayes feature set, categorical features (balls/strikes as categories) + Gaussian numerics
nb_vars <- c("pitch_name", "balls", "strikes", "stand", "p_throws",
             "release_speed", "release_spin_rate", "pfx_x", "pfx_z")
nb_train <- training_df %>% mutate(balls = factor(balls), strikes = factor(strikes))
nb_test  <- testing_df  %>% mutate(balls = factor(balls, levels = levels(nb_train$balls)),
                                   strikes = factor(strikes, levels = levels(nb_train$strikes)))

# Zero-frequency check, empty (feature level x class) cells in the training data
nb_cat_vars <- c("pitch_name", "balls", "strikes", "stand", "p_throws")
zero_check <- data.frame(
  Feature = nb_cat_vars,
  Zero_cells = sapply(nb_cat_vars, function(v) sum(table(nb_train[[v]], nb_train$Y) == 0)))

# Conditional-independence check, correlation among the numeric predictors
num_cor <- round(cor(training_df[, c("release_speed", "release_spin_rate", "pfx_x", "pfx_z")]), 2)

# Fit Naive Bayes with Laplace smoothing
nb_fit <- naiveBayes(Y ~ ., data = nb_train[, c("Y", nb_vars)], laplace = 1)
P_nb   <- predict(nb_fit, nb_test[, nb_vars], type = "raw")   # soft posteriors, columns = Ball, Strike, InPlay
colnames(P_nb) <- classes
```

```{r, echo=FALSE, fig.height=4, fig.width=7, fig.align="center"}
# Density Plots 
training_df %>%
  select(Y, release_speed, release_spin_rate, pfx_x, pfx_z) %>%
  pivot_longer(-Y) %>%
  ggplot(aes(value, colour = Y)) + geom_density() +
  facet_wrap(~ name, scales = "free") + 
  labs(x = NULL, title = "Density Plots of Numerical Predictors",
       colour = "Pitch result")
```

Naive Bayes assumes the features are conditionally independent given the pitch result. Numeric features (velocity, spin rate, horizontal and vertical break) follow Gaussian class-conditional densities. Categorical features (pitch type, balls, strikes, batter side, pitcher hand) use class-conditional frequency tables.

In the density plots above, each show one numeric predictor, with one curve representing one pitch result. In every density plot, all lines almost completely overlap in every panel. If a ball, a strike, and ball put in play pitch all have nearly the same velocity, spin, and break distributions, those features can barely tell the classes apart. 

Another thing to note is that the density curves are not Gaussian, or Normal. Vertical break (pfx_z) and release speed (release_speed) are bimodal and horizontal break (pfx_x) is pretty bumpy. The best distribution is release spin rate (release_spin_rate), which has a dominant peak, but is not completely smooth.

```{r, echo=FALSE}
# Correlation table
kable(num_cor, caption = "Correlation among numeric predictors.")
```

Looking at the correlation table, most pairs are close to zero. The exception is release_speed and pfx_z, which have a value of 0.72. This is likely because faster pitches tend to have more vertical movement. Since Naive Bayes treats the two features as independent evidence, it essentially counts the same information twice. That tends to make its probabilities more extreme than they should be, which the calibration plot should show.

Given the information above, we suggest that there is not independence within the data. 

**Zero-frequency problem:** 

```{r, echo=FALSE}
kable(zero_check, row.names = FALSE, caption = "Zero-frequency check, empty feature-level-by-class cells.")
```

Every categorical feature (pitch type, balls, strikes, batter side, pitcher hand) was checked for feature-level-by-class combinations with no training observations, and found none. Regardless, we used Laplace smoothing (laplace = 1) so that a feature level never seen with a class in training cannot force that class's posterior to exactly zero. We extracted both the hard predicted classes (argmax of the posterior) and the soft posteriors $\hat{P}(Y = k \mid X)$.

**Hard Predicted Classes and Soft Posteriors:** 
```{r, echo = FALSE}
# Hard classes and soft posteriors for Naive Bayes
pred_nb <- max.col(P_nb, ties.method = "first")

# First 6 test pitches
post_tab <- data.frame(Actual = testing_df$Y, round(P_nb, 3), Predicted = classes[pred_nb])[1:6, ]
kable(post_tab, row.names = FALSE,
      caption = "Naive Bayes soft posteriors $\\widehat{P}(Y = k \\mid X)$ and hard predicted class for the first six test pitches.")
```

**Confusion Matrix:** 
```{r, echo = FALSE}
# Naive Bayes confusion matrix (test set); rows = predicted, columns = actual
cm_nb <- table(predicted = factor(classes[pred_nb], levels = classes),
               actual = factor(classes[as.integer(testing_df$Y)], levels = classes))
kable(cm_nb, caption ="Naive Bayes confusion matrix (test set). Rows = predicted, columns = actual.")
```


# 4. Analysis, Tuning & Results
<!-- 
* NEED TO DO THIS: Include tuning plots and tables for the optimal k selection (regression and classification).
* Provide the comparative evaluation table of classification metrics described in Section 4 (both models plus the no-skill baseline), with a sentence justifying your choice of metrics.
* Include at least one visualization contrasting the behavior of Naive Bayes and kNN, e.g., ROC curves overlaid on the same axes (binary), a calibration plot, or a 2D decision boundary plot.
---> 

**[NEED TO DO THIS: Include tuning plots and tables for the optimal k selection (regression and classification).]**

## 4.1: Regression (kNN Regression)
**Root Mean Squared Error (RMSE):**
**Mean Absolute Error (MAE):**
**Coefficient of Determination** ($R^2$)**:**

## 4.2 Classification (kNN & Naive Bayes)
```{r, echo=FALSE}
P_null <- matrix(as.numeric(pi_train), nrow = nrow(testing_df), ncol = K,
                    byrow = TRUE, dimnames = list(NULL, classes))
pred_null <- rep(which.max(pi_train), length(yte))

# Evaluate one model on the test set (multi-class)
eval_model <- function(P, y, pred) {
  n <- length(y)
  tab <- table(Pred = factor(pred, 1:K), Actual = factor(y, 1:K))
  tp <- diag(tab);  rs <- rowSums(tab);  cs <- colSums(tab)
  rec <- tp / cs
  prec <- ifelse(rs > 0, tp / rs, 0)
  f1 <- ifelse(prec + rec > 0, 2 * prec * rec / (prec + rec), 0)
  po <- sum(tp) / n;  pe <- sum(rs * cs) / n^2
  den <- sqrt((n^2 - sum(rs^2)) * (n^2 - sum(cs^2)))
  Y1 <- diag(K)[y, ]
  c(Accuracy = po,
    BalancedAcc = mean(rec),
    MacroF1 = mean(f1),
    MCC = ifelse(den > 0, (sum(tp) * n - sum(rs * cs)) / den, 0),
    Kappa = (po - pe) / (1 - pe),
    LogLoss = multi_log_loss(y, P),
    Brier = mean(rowSums((P - Y1)^2))) 
}

res <- rbind("No-skill baseline" = eval_model(P_null, yte, pred_null),
             "kNN" = eval_model(P_knn,  yte, pred_knn),
             "Naive Bayes" = eval_model(P_nb,   yte, pred_nb))
res <- cbind(res, LL_Improvement = 1 - res[, "LogLoss"] / res["No-skill baseline", "LogLoss"])
```

**Confusion Matrices**
```{r, echo=FALSE,fig.align="center", results='asis'}
cm_knn <- table(predicted = factor(classes[pred_knn], levels = classes),
                actual = factor(classes[yte], levels = classes))
cm_nb <- table(predicted = factor(classes[pred_nb], levels = classes),
                actual = factor(classes[yte], levels = classes))

kable(cm_knn, caption = paste0("kNN (k* = ", k_star, ") confusion matrix. Rows = predicted, columns = actual."))
kable(cm_nb, caption = "Naive Bayes confusion matrix. Rows = predicted, columns = actual.")
```

Looking at both confusion matrices of kNN and Naive Bayes, both models most accurately predict Strike. For kNN, accuracy is 0.469, which is essentially the same as always guessing Strike (0.4667). Therefore, accuracy here is misleading. Its accuracy matches the baseline only because it predicts the majority class almost every time. 

A worthy note is that kNN does not predict InPlay at all. This is because of the large $k^*$ (201). While Naive Bayes was able to predict all three classes, it only predicted InPlay 63 times, and mostly wrong. 

```{r, echo =FALSE}
# comparative evaluation table
comp <- t(res[, c("Accuracy", "BalancedAcc", "MCC", "LogLoss", "LL_Improvement", "Brier")])
rownames(comp) <- c("Accuracy", "Balanced accuracy", "MCC", "Log-loss",
                    "Log-loss improvement over null", "Brier score")

# Which model is better on each metric, lower is better for log-loss and Brier
lower_better <- c(FALSE, FALSE, FALSE, TRUE, FALSE, TRUE)
two <- comp[, c("kNN", "Naive Bayes")]
best_idx <- ifelse(lower_better, apply(two, 1, which.min), apply(two, 1, which.max))

comp_tab <- data.frame(Metric = rownames(comp), round(comp, 4),
                       `Better model` = colnames(two)[best_idx],
                       check.names = FALSE, row.names = NULL)

kable(comp_tab,
      caption = "Comparative test-set evaluation.")
```

No-skill baselines: accuracy = No Information Rate, balanced accuracy = 1/3, MCC = 0, Brier = $1 - \sum_k \pi_k^2$.

**Log-loss from the predicted probabilites**
The table above represents the log-loss from the predicted probabilities. For log-loss, the lower the value the better. When compared to the null baseline model, kNN produces a smaller log-loss value while naive bayes grew. Therefore, by this metric, we suggest kNN is better.

**Null Baseline Entropy**
The table above represents null baseline entropy. From the table above, kNN make a 1.0133 improvement from the null model, while Naive Bayes decreases by 1.7. Therefore, by this metric, we suggest kNN is better.

**Balanced Accuracy**
The table above represents the balanced accuracy of kNN and Naive Bayes. For balanced accuracy, the higher the number the better. kNN had a positive increased of 0.0137. On the other hand, Naive Bayes had a slightly bigger positive increase of 0.0144. Therefore, based off this metric, we can conclude that Naive Bayes is better.

**Matthews Correlation Coefficient**
The table above represent the MCC (Matthews Correlation Coefficient) of kNN and Naive Bayes. MCC prefers a higher number. Looking at the table above, kNN had a positive difference of 0.0427 while Naive Bayes had a positive, but not as positive, difference of 0.0327. Therefore, based off this metric, kNN is better. 

**Brier Score**
The table above represents the multi-class Brier score. We want the Brier score to be as low as possible, so ideally we want kNN and Naive Bayes to have a negative difference from the baseline. Looking at the table above, kNN barely changes from the null model and Naive Bayes make a positive difference Therefore, the no-skill model is the best. 

**Calibration Plots**
```{r, echo=FALSE, fig.width=7, fig.height=3.5, out.width="90%", fig.align="center"}
# Calibration Plot 
calib_df <- function(P, y, model) {
  do.call(rbind, lapply(1:K, function(j) {
    d <- data.frame(p = P[, j], obs = as.integer(y == j))
    d$bin <- cut(d$p, breaks = seq(0, 1, by = 0.05), include.lowest = TRUE)
    d %>% group_by(bin) %>%
      summarise(mean_pred = mean(p), obs_freq = mean(obs), n = n(), .groups = "drop") %>%
      filter(n >= 20) %>% 
      mutate(class = classes[j], model = model)
  }))
}

cal <- rbind(calib_df(P_knn, yte, paste0("kNN")),
             calib_df(P_nb,  yte, "Naive Bayes"))
cal$class <- factor(cal$class, levels = classes)

ggplot(cal, aes(mean_pred, obs_freq, colour = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_line() + geom_point(aes(size = n), alpha = 0.8) +
  facet_wrap(~ class) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Mean predicted probability", y = "Observed frequency", 
       title = "Calibration Plot of kNN and Naive Bayes",
       colour = NULL, size = "Pitches") +
  theme(legend.position = "bottom")
```

The plots above represent the calibration plots, showing whether the predicted probability really corresponds to that class happening. 

For Ball, Naive Bayes predicts about 0.68 at its upper end, but is has an observed frequency of roughly 0.49. In this case, Naive Bayes is over-confident kNN is better, with its upper end matching much closer to the observed frequency. 

For Strike, both kNN and Naive Bayes have much flatter lines than the observed frequency line. In other words, the predictions are too extreme in both directions. On the lower end of Naive Bayes, the predicted probability is much higher than the observed diagonal. 0.8 predicted probability compared to a .30 observed frequency. In this case, Naive Bayes is under-confident. 

For InPlay, the lines follow the observed diagonal the best out of all the other classes. Although, on the upper end of Naive Bayes, the model appears to be slighly over-confident. 

In general, the kNN models better follows the observed frequency diagonal. The Naive Bayes model tends to wonder off towards then end of the lines. With this being said, for both models, most observations are concentrated in the middle of the model. 

# 5. Discussion & Real-World Implications
<!-- 
address the discussion prompts listed in Section 6.
--->
**Accuracy vs. Log-Loss Discrepancies**

**Generative vs. non-parametric trade-offs**

**Neighborhood size behavior in kNN**

**Real-world decision context**


# 6. Citations & Data Link
<!-- 
link to the original data source and provide a direct, one-click access link to your data file (e.g., hosted on Google Drive).
--->
Baseball Savant Website: [Original data source](https://baseballsavant.mlb.com/statcast_search)

Data File: [Clean Data](https://docs.google.com/spreadsheets/d/1_zxPBT3iTcbFs3PdsiJ9JdB-IF8XxnXb9x7-mVuE-XU/edit?usp=sharing)

# 7. Appendix
<!-- 
all clean, commented R code necessary to fully reproduce your training splits, models, metric calculations, tables, and figures.
--->










